---
title: "GRASP and ILS for the flowshop problem"
author: "Jose M Sallan"
date: "01/04/2020"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The GRASP function

The GRASP strategy consists of building starting solutions for the local search algorithm based on a constructive heuristic. The difference with the constructive heuristic is that instead of picking the best element according with the algorithm, we pick one element at random from a **restricted category list**. The number of elements of this list is the parameter `rcl` of the algorithm.

For the permutative flowshop, the `PalmerGenerator` generates a random solution based on the Palmer heuristic. This heuristic orders the jobs by increasing order of $S_1 - S_2$:

\[
S_1 = \displaystyle\sum_{i=1}^n \left(m - i\right)t_{ij}
\]

\[
S_2 = \displaystyle\sum_{i=1}^n \left(i - 1\right)t_{ij}
\]

`PalmerGenerator` constructs the solution by picking not the element available of minimum value of $S_1 - S_2$ at each step, but one of the best `rcl` elements available.

```{r, eval=FALSE}
# generator of starting solutions for a GRASP heuristic
PalmerGenerator <- function(M, rcl=4){
  
  m <- dim(M)[1]
  n <- dim(M)[2]
  S <- matrix(0, 2, n)
  
  for(j in 1:n)
    for(i in 1:m){
      S[1, j] <- S[1, j] + (m - i)*M[i, j]
      S[2, j] <- S[2, j] + (i - 1)*M[i, j]
    }
  
  
  diff <- S[1, ] - S[2, ]
  palmer <- order(diff)
  
  grasp <- numeric(n)
  
  chosen <- rep(FALSE, n)
  
  for(i in 1:n){
    k <- sample(1:min(rcl, n-i+1), 1)
    j <- which(chosen==FALSE)[k]
    grasp[i] <- palmer[j]
    chosen[j] <- TRUE
  }
  
  return(grasp)
}
```

`GRASPFS` is a implementation of GRASP for the permutative flowshop problem. It uses `PalmerGenerator` for generating starting solutions, and hill climbing (HC), simulated annealing (SA) and tabu search (TS) as local search algorithms. Note the use of the ellipsis `...` parameter in the function to pass additional parameters to the simulated annealing function. You can learn more about the ellipsis here:

https://www.r-bloggers.com/r-three-dots-ellipsis/

The `iter` parameter controls the number of iterations to run.

```{r, eval=FALSE}
GRASPFS <- function(M, rcl=4, iter=100, op="swap", opt="HC", ...){

  params <- list(...)

  if(!opt %in% c("HC", "TS", "SA"))
    return(success==FALSE)

  n <- dim(M)[2]
  bestsol <- numeric(n)
  bestfit <- Inf

  for(t in 1:iter){

    seed_sol <- PalmerGenerator(M, rcl=rcl)

    if(opt=="HC")
      test_sol <- HCFS(M, seed_sol, op=op)

    if(opt=="TS")
      test_sol <- TSFS(M, seed_sol, iter=25, op=op, early = TRUE)

    if(opt=="SA"){
      if(is.null(params$Tmax) & is.null(params$mu))
        test_sol <- SAFS(M, seed_sol, op=op)
      if(!is.null(params$Tmax) & is.null(params$mu))
        test_sol <- SAFS(M, seed_sol, Tmax=params$Tmax, op=op)
      if(is.null(params$Tmax) & !is.null(params$mu))
        test_sol <- SAFS(M, seed_sol, mu=params$mu, op=op)
      if(!is.null(params$Tmax) & !is.null(params$mu))
        test_sol <- SAFS(M, seed_sol, Tmax=params$Tmax, mu=params$mu, op=op)
    }


    if(test_sol$obj < bestfit){
      bestsol <- test_sol$sol
      bestfit <- test_sol$obj
    }

  }

  return(list(sol=bestsol, obj=bestfit))
}
```


# The ILS function

The ILS strategy consists of getting the result of the local search algorithm, submitting it to a perturbation and use the resulting function as a new starting solution. The perturbation we have chosen is to perform several random insertions. The number of insertions `ni` to perform is a parameter of the algorithm.

```{r, eval=FALSE}
PerturbationInsertion <- function(v, ni=4){
  n <- length(v)
  for(i in 1:ni){
    ch <- sample(1:n, 2)
    v <- insertion(v, ch[1], ch[2])
  }
  return(v)
}
```

`ILFS` is an implementation of iterative local search for the flowshop problem. The solution to perturbate in each step sol is replaced by the result of local search test_sol only if the later has a lower makespan than the former. With `bestsol` we keep track of the best result we have found along the way. Similarly as `GRASPFS`, the `iter` parameter controls the number of iterations to run.

```{r, eval=FALSE}
ILSFS <- function(M, ni=4, iter=100, opt="HC", ...){

  params <- list(...)

  if(!opt %in% c("HC", "TS", "SA"))
    return(success==FALSE)

  n <- dim(M)[2]
  bestsol <- numeric(n)
  bestfit <- Inf

  sol <- PalmerTrapezes(M)$pal
  fit <- makespan(M, sol)

  for(i in 1:iter){

    seed_sol <- PerturbationInsertion(sol)

    if(opt=="HC")
      test_sol <- HCFS(M, seed_sol, op=op)

    if(opt=="TS")
      test_sol <- TSFS(M, seed_sol, iter=25, op="swap", early = TRUE)

    if(opt=="SA"){
      if(is.null(params$Tmax) & is.null(params$mu))
        test_sol <- SAFS(M, seed_sol, op="swap")
      if(!is.null(params$Tmax) & is.null(params$mu))
        test_sol <- SAFS(M, seed_sol, Tmax=params$Tmax, op="swap")
      if(is.null(params$Tmax) & !is.null(params$mu))
        test_sol <- SAFS(M, seed_sol, mu=params$mu, op="swap")
      if(!is.null(params$Tmax) & !is.null(params$mu))
        test_sol <- SAFS(M, seed_sol, Tmax=params$Tmax, mu=params$mu, op="swap")
    }

    if(test_sol$obj < fit){
      sol <- test_sol$sol
      fit <- test_sol$obj
    }

    if(test_sol$obj < bestfit){
      bestsol <- test_sol$sol
      bestfit <- test_sol$obj
    }
  }

  return(list(sol=bestsol, obj=bestfit))

}
```

# Testing the functions

To examine the performance of the functions, we use its implementation on the combheuristics package. For accessing the package, see:

https://github.com/jmsallan/combheuristics 

```{r}
library(combheuristics)
```


To test how do these functions perform, we can pick a sample instance:

```{r}
set.seed(2020)
instance <- matrix(sample(10:90, 100, replace = TRUE), 5, 20)
```

Here is the code evaluating `GRASPFS`:

```{r}
set.seed(2020)

#Standard options of the GRASP function with iter=50
GRASP01 <- GRASPFS(instance, iter=50)
GRASP01$obj

#GRASP with simulated annealing.
GRASP02 <- GRASPFS(instance, opt="SA", iter=50, Tmax=5000, mu=2500)
GRASP02$obj

#GRASP with tabu search.
GRASP03 <- GRASPFS(instance, iter=50, opt="TS")
GRASP03$obj
```

And here is the code evaluating `ILSFS`:

```{r}
#Standard options of the ILS function, including Hill climbing.
ILS01 <- ILSFS(instance, iter=50)
ILS01$obj

#GRASP with simulated annealing.
set.seed(2020)
ILS02 <- ILSFS(instance, iter=50, opt="SA", Tmax=5000, mu=2500)
ILS02$obj

#ILS with tabu search.
ILS03 <- ILSFS(instance, iter=50, opt="TS")
ILS03$obj
ILS03$sol
```


