---
title: "Speed of for loops"
author: "Jose M Sallan"
date: "05/03/2020"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Our aim is to sum the columns of a very large matrix. We have three ways of doing that:

To defined a function that sums each column and puts the value into a vector component using a for loop:

```{r}
f1 <- function(m){
  n <- dim(m)[2]
  s <- numeric(n)
  for(i in 1:n) s[i] <- sum(m[,i])
  return(s)
}
```

Perform the sum of columns with an `apply` loop over columns:

```{r}
f2 <- function(m) return(apply(m, 2, sum))
```

Use the base function `colSums`.

Let's define a very large matrix:

```{r}
M <- matrix(sample(1:100, 1000000, replace = TRUE), 1000, 1000)
```

and apply each function to it:

```{r}
s1 <- f1(M)
s2 <- f2(M)
s3 <- colSums(M)
```

The `s2` vector is integer, and the other two numeric. Let's check if they yield the same values:

```{r}
identical(as.integer(s1), s2)
identical(s1, s3)
identical(s2, as.integer(s3))
```

Let's check the speed of each function:

```{r}
library(rbenchmark)
benchmark(f1(M), f2(M), colSums(M), columns=c('test', 'replications', 'elapsed', 'relative', 'user.self', 'sys.self'), order='elapsed')
```

The best performance is achieved by the built-in function. The `for` loop implementation seems to go faster than `apply` in this context.
