---
title: "Three insertion operators"
author: "Jose M Sallan"
date: "3/17/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Three insertion operators

Let's compare three implementations of an `insertion` function which inserts element `i` of a vector in position `j`. These three operators came from responses of students required to code this operator.

```{r}
insertion1 <- function(v, i, j){
  if(i < j){
    aux <- v[i]
    for(i in i:(j-1)){
      v[i] <- v[i+1]
    }
    v[j] <- aux
    return(v)
  }
  else if(i > j){
    aux <- v[i]
    for(i in i:(j+1)){
      v[i] <- v[i-1]
    }
    v[j] <- aux
    return(v)
  }
  else return(v)
}

insertion2 <- function(v, i, j){
  
  if(i==j) return(v)
  aux<- v[i]
  
  if(i<j){ v[i:(j-1)] <- v[(i+1):j] }
  if(i>j){ v[(j+1):i] <- v[j:(i-1)] }
  v[j] <- aux
  return(v)
}

insertion3 <- function(v, i, j){
  
  aux <- v[i]
  v<-append(v[-i],aux,j-1)
  
  return(v)
}
```

Let's examine the three proposals:

* `insertion1` is coded in a similar way as in C++ or Java, processing the function component by component. To code the function in this way, we need to use for loops.
* `insertion2` is the "official" answer to the coding, which takes advantage of the vectorial subsetting capabilities of R.
* `insertion3` uses a built-in append function to do the insertion.

Let's build a test sample:

```{r}
set.seed(2020)
test <- sample(1:5, 5)
test
```

Let's enlist all possible insertions for this sample for the three operators:

```{r}
list_insertion1 <- list()
list_insertion2 <- list()
list_insertion3 <- list()
n <- length(test)
k <- 1
for(i in 1:n)
  for(j in 1:n)
    if(i!=j & i!=(j-1)){
      list_insertion1[[k]] <- insertion1(test, i, j)
      list_insertion2[[k]] <- insertion2(test, i, j)
      list_insertion3[[k]] <- insertion3(test, i, j)
      k <- k+1
    }
```

Let's see how it works:

```{r}
test
list_insertion1
```

And if all the three are the same:

```{r}
identical(list_insertion1, list_insertion2)
identical(list_insertion1, list_insertion3)
```


Let's build functions that compute all the insertions of a vector of length `n` with each insertion function:

```{r}
allinsertions1 <- function(n){
  test <- 1:n
  l <- list()
  k <- 1
for(i in 1:n)
  for(j in 1:n)
    if(i!=j & i!=(j-1)){
      l[[k]] <- insertion1(test, i, j)
      k <- k+1
    }
  return(l)
}

allinsertions2 <- function(n){
  test <- 1:n
  l <- list()
  k <- 1
for(i in 1:n)
  for(j in 1:n)
    if(i!=j & i!=(j-1)){
      l[[k]] <- insertion2(test, i, j)
      k <- k+1
    }
  return(l)
}

allinsertions3 <- function(n){
  test <- 1:n
  l <- list()
  k <- 1
for(i in 1:n)
  for(j in 1:n)
    if(i!=j & i!=(j-1)){
      l[[k]] <- insertion3(test, i, j)
      k <- k+1
    }
  return(l)
}

```

And now let's compare performance of each function for a short vector:

```{r}
library(rbenchmark)
n <- 10
benchmark(allinsertions1(n), allinsertions2(n), allinsertions3(n), columns=c('test', 'replications', 'elapsed', 'relative', 'user.self', 'sys.self'), replications=100, order='elapsed')
```

Now with a longer vector:

```{r}
library(rbenchmark)
n <- 50
benchmark(allinsertions1(n), allinsertions2(n), allinsertions3(n), columns=c('test', 'replications', 'elapsed', 'relative', 'user.self', 'sys.self'), replications=100, order='elapsed')
```

And with an even longer vector:

```{r}
library(rbenchmark)
n <- 100
benchmark(allinsertions1(n), allinsertions2(n), allinsertions3(n), columns=c('test', 'replications', 'elapsed', 'relative', 'user.self', 'sys.self'), replications=100, order='elapsed')
```

As a conclusion, the less effective implementation is the `insertion3` function. `insertion1` works better for small vectors, although `insertion2` works much better for longer vectors.
