---
title: "Solving flowshop with the GA package"
author: "Jose M Sallan"
date: "01/04/2020"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The GA package for genetic algorithms

We will use the `GA` package, written by Luca Scrocca to implement
genetic algorithms in the R environment. A useful vignette can be found in:

https://cran.r-project.org/web/packages/GA/vignettes/GA.html

```{r, message=FALSE}
library(GA)
```


# Using GA to optimize makespan in flowshop

The vignette is focused in real-value and binary encodings. I will show here some additional information about how to use permutation encodings using the flowshop as benchmark.

First, we must define a makespan function adequate for GA. As GA solver are maximizers, the function will return the value of makespan multiplied by -1.

```{r}
makespanGA <- function(M, sol){
  
  m <- dim(M)[1]
  n <- dim(M)[2]
  
  M <- M[ , sol]
  
  for(j in 2:n) M[1, j] <- M[1, (j-1)] + M[1, j]
  for(i in 2:m) M[i, 1] <- M[(i-1), 1] + M[i, 1]
  
  for(i in 2:m)
    for(j in 2:n)
      M[i,j] <- max(M[i-1,j], M[i, j-1]) + M[i, j]
  
  
  return(-M[m, n])
}
```

To test the function, we will use the same instance that we have used for local search algorithms:

```{r}
set.seed(2020)
instance <- matrix(sample(10:90, 100, replace = TRUE), 5, 20)
```

Now let's solve the GA with the function defaults. We will make the following choices here:

* The solutions are of the `permutation` type.
* The lower and upper parameters correspond with the lowest and highest value that can take the permutation. In this case, 'lower=1' and `upper` is equal to the number of columns of the instance matrix.
* we will hand to the function the parameter `M` of the `makepanGA` function.
* We will add a `seed` value for reproducibility, as the genetic algorithm includes randomness in their operators.

```{r}
ga01 <- ga(type = "permutation", fitness = makespanGA, M=instance, lower = 1, upper = dim(instance)[2], seed=1111)
```

Let's see the solution value and the value of fitness function (note that we have to change the sign of the fitness value function):

```{r}
ga01@solution
-ga01@fitnessValue
```

We can make a plot of the evolution of the algorithm:

```{r, fig.width=11}
plot(ga01)
```

# Tuning the parameters

Let's take a look at the parameter defaults:

```{r}
ga01@popSize #population size
ga01@maxiter #number of generations
ga01@pcrossover #probability of crossover between pairs of chromosomes
ga01@pmutation #the probability of mutation in a parent chromosome
```

We can check what happens if we tune these parameters:

```{r}
ga02 <- ga(type = "permutation", fitness = makespanGA, M=instance, lower = 1, upper = dim(instance)[2],  popSize = 100, pcrossover = 1, maxiter = 300, seed=1111)
-ga02@fitnessValue
```

Let's see algorithm evolution:

```{r}
plot(ga02)
```


# Selection, crossover and mutation operators

The algorithm operators can be retrieved looking at the `gaControl` function:

```{r}
gaControl("permutation")
```

To see the possible values of operators we can look at the article presenting the algorithm, available at the website of the *Journal of Statistical Software*:

https://www.jstatsoft.org/article/view/v053i04

The available operators for permutation listed in the article are:

Selection operators:

* Linear-rank selection (default): `gaperm_lrSelection()`
* Nonlinear-rank selection: `gaperm_nlrSelection()`
* Proportional (roulette-wheel) selection: `gaperm_rwSelection()`
* Unibaised tournament selection: `gaperm_tourSelection()`

Crossover operators:

* Cycle crossover: `gaperm_cxCrossover()`
* Partially matched crossover: `gaperm_pmxCrossover()`
* Order crossover (default): `gaperm_oxCrossover()`
* Position-based crossover: `gaperm_pbxCrossover()`

Mutation operators:

* Simple inversion mutation (default): `gaperm_simMutation()` DEFAULT
* Insertion mutation: `gaperm_ismMutation()`
* Swap mutation: `gaperm_swMutation()`
* Displacement mutation: `gaperm_dmMutation()`
* Scramble mutation: `gaperm_scrMutation()`

For a reference for the meaning of the existing operators, you can check:

http://www.rubicite.com/Tutorials/GeneticAlgorithms.aspx

We can change the default operators with the `gaControl` function:

```{r}
#saving, just in case, the default parameters:
defaultControl <- gaControl("permutation")

#changing operators:
gaControl("permutation" = list(selection = "gaperm_nlrSelection", crossover = "gaperm_cxCrossover", mutation = "gaperm_ismMutation"))
```

Then, we can check algorithm performance with the new operators:

```{r}
ga03 <- ga(type = "permutation", fitness = makespanGA, M=instance, lower = 1, upper = dim(instance)[2],  popSize = 100, pcrossover = 1, maxiter = 300, seed=1111)
-ga03@fitnessValue
```

Let's check another set of parameters:


```{r}
ga04 <- ga(type = "permutation", fitness = makespanGA, M=instance, lower = 1, upper = dim(instance)[2], popSize=100, maxiter = 100, pmutation = 1, seed=1111)
-ga04@fitnessValue
```

This version of the GA is the best available for this instance. In fact, it has found a better solution than the obtained with local search algorithms in only 100 iterations:

```{r}
plot(ga04)
```


Let's show the solution for the record:

```{r}
ga04@solution
```


